\section{Machine Learning Fundamentals}\label{sec:bg:ml}

At its most general, \gls{ml} replaces hand-coded decision rules with parametric models whose parameters are fitted to data so that a chosen performance criterion is optimized on observed examples of the task.
A model is a function \(\Enc\colon \setX \to \setY\) from an input space \(\setX\) to a task-specific output space \(\setY\), and learning consists of searching over parameter vectors \(\bm{\theta}\) until \(\Enc\) is consistent with the regularities present in the training data.
What changes between problems is the nature of the supervision available, the structure of \(\setY\), and the loss that scores predictions against that supervision.

A common formal frame is \emph{empirical risk minimization}.
Given a dataset \(\setD = \{(\vect{x}_i, y_i)\}_{i=1}^{N}\) drawn from an unknown distribution and a pointwise loss \(\ell\), the learner solves
\begin{equation}\label{eq:bg:erm}
  \hat{\bm{\theta}}
  =
  \argmin_{\bm{\theta}}\;
  \frac{1}{N}\sum_{i=1}^{N} \ell\bigl(\Enc(\vect{x}_i),\, y_i\bigr) + \Omega(\bm{\theta}),
\end{equation}
where \(\Omega\) is an optional regularizer that trades fidelity to the training data against complexity of the fitted model.
Generalization to fresh inputs is then assessed on held-out data, and the entire training procedure---data, model class, loss, optimizer, regularization---is judged by its effect on that out-of-sample behavior rather than on the training fit alone.

\subsection{Learning paradigms}

\Gls{ml} problems are conventionally separated by the kind of supervision available at training time.
In \emph{supervised} learning, each \(\vect{x}_i\) is paired with a label \(y_i\), categorical for classification or real-valued for regression, and \(\ell\) penalizes disagreement between the prediction and that label.
In \emph{unsupervised} learning, labels are absent and the goal is to expose structure in the input distribution itself, for example through density estimation, dimensionality reduction, or clustering; \cref{sec:bg:clustering} treats clustering specifically because it is the central downstream consumer of the embeddings used in this thesis.
\emph{Self-supervised} learning sits between the two: it constructs supervisory targets automatically from the data through a designed pretext task---predicting masked parts of an input, contrasting augmented views, or ordering shuffled tokens---so that an encoder can be trained without manual annotation.
The remainder of this thesis combines an unsupervised clustering objective with a self-supervised pretext signal; the semi-supervised setting, in which a small labeled subset is augmented with abundant unlabeled data, is not considered.

\subsection{Representation learning}

Much of modern \gls{dl} practice is best understood as \emph{representation learning}: a parametric encoder maps high-dimensional observations into a lower-dimensional vector space whose geometry is shaped by the training objective.
Downstream tasks---classification, clustering, retrieval, or sequence prediction---then operate on the resulting \emph{embeddings} \(\vect{z} = \Enc(\vect{x}) \in \R^{d}\) rather than on raw sensor coordinates.
The advantage of routing tasks through a learned representation is that invariances, similarities, and task-relevant factors of variation can be encoded by \(\Enc\) once, after which simple linear or geometric operations on \(\vect{z}\) suffice for many downstream questions.
\Cref{sec:bg:dl,sec:bg:transformer} expand on this view; the method chapter (\cref{ch:method}) instantiates it for \gls{pdw} streams, training an encoder whose embeddings are intended to expose emitter and mode structure.

\subsection{Applications in radar deinterleaving and mode classification}

Radar signal deinterleaving and operating-mode classification illustrate the practical pull of these ideas.
For decades, both tasks were addressed by deterministic temporal rules and hand-crafted statistics over \gls{pdw} parameters, as outlined in \cref{sec:bg:elint_pipeline}.
As emitters became more agile and electromagnetic environments more crowded, the field migrated first to \emph{shallow} \gls{ml} in the form of multi-parameter clustering---\(k\)-means and density-based methods over \gls{rf}, \gls{aoa}, and pulse width---and more recently to \emph{deep} sequence models that learn pulse-level features directly from data \parencite{qu2026radardeinterleavingreview}.
Recurrent and attention-based architectures now treat deinterleaving as sequence labeling in which the per-pulse output identifies an emitter, while transformer encoders in particular are well suited to capturing the non-contiguous, multi-scale relationships induced by jittered, staggered, or grouped \glspl{pri} \parencite{qu2026radardeinterleavingreview,gunn2025deinterleaving}.

Mode classification follows a parallel trajectory.
Classical \gls{esm} systems match deinterleaved pulse trains against catalogued waveform templates in a mode library \parencite{wiley2006elint}, while learning-based approaches reframe the same task either as supervised classification over known mode labels or as unsupervised grouping over pulse trains when labels are scarce or incomplete \parencite{qu2026radardeinterleavingreview}.
The latter setting is the one this thesis adopts: emitter and mode structure is to be \emph{discovered} jointly from the input distribution, rather than recognized against a fixed catalogue.
The clustering tools in \cref{sec:bg:clustering} and the representation learning machinery in \cref{sec:bg:dl,sec:bg:transformer} provide the technical foundation that the method chapter builds on.
