\section{Problem Formulation}\label{sec:method:formulation}

This chapter studies supervised representation learning from time-ordered streams of \glspl{pdw} produced by an \gls{esm} or deinterleaving front end, as outlined in \cref{sec:bg:elint_pipeline}.
Each pulse in the training data carries two ground-truth labels: an \emph{emitter identity}, whose values are unique only within the pulse train that contains the pulse, and an \emph{operating mode}, whose label set is a finite catalogue shared across all pulse trains.
Both labels are used to supervise embedding geometry during training, and both per-pulse outputs are produced by clustering the resulting embeddings at inference time.
The goal is twofold: learn a representation from which (i) pulses originating from the same physical emitter within a pulse train can be reliably grouped by clustering the emitter embeddings, and (ii) pulses sharing the same operating mode --- including modulation types not present in the training catalogue --- can be reliably grouped by clustering the mode embeddings.
Routing the mode output through clustering rather than through a closed-catalogue classifier is a deliberate design choice: it keeps the system open to previously unseen modulation variants, an experimental scenario investigated in \cref{ch:experiments} by holding out selected modulations during training and verifying that the model still groups their pulses into coherent mode clusters.


\subsection*{Observed input}

The data is organised as a collection of \emph{pulse trains}, each a finite, time-ordered sequence of \glspl{pdw} produced during one observation window.
A single pulse train of $N$ pulses is represented by feature vectors
\begin{equation}\label{eq:method:pdw_sequence}
  \mat{X} = (\vect{x}_1,\ldots,\vect{x}_N),
  \qquad
  \vect{x}_n \in \R^{d},
\end{equation}
where each $\vect{x}_n$ encodes the measured pulse parameters associated with one \gls{pdw} (carrier frequency, amplitude, pulse width, time of arrival, angle of arrival including azimuth and elevation).
The model receives one pulse train at a time and produces per-pulse outputs for that train; to keep notation light, we suppress an explicit train index in \cref{eq:method:pdw_sequence} and in what follows.

A single train may contain irregular sampling, missing pulses, and spurious detections inherited from upstream processing; robustness to such effects is a design constraint rather than part of the formal specification.

\subsection*{Per-pulse supervision: emitter and mode labels}

Every pulse in the training data is annotated with two ground-truth labels.
For each index $n \in \{1,\ldots,N\}$ within a given train, let $e_n$ denote the emitter identity responsible for pulse $n$ and let $m_n$ denote its operating mode.
Although both quantities are observed during training, they have fundamentally different semantics.

The mode label $m_n \in \{1,\ldots,M\}$ is drawn from a finite, \emph{global} catalogue of size $M$ that is shared across all pulse trains: the symbol $m_n = \ell$ denotes the same operating mode regardless of which train the pulse belongs to.
The emitter label $e_n \in \{1,\ldots,K_{\mathrm{em}}\}$, by contrast, is unique only \emph{within the current train}: the integer indexes the distinct physical emitters encountered in this particular observation window, and the same integer in a different train refers to an unrelated emitter.
Consequently, two pulses can be declared to share an emitter only when both belong to the same train, and the supervision provides no notion of a global emitter identity.

The train-specific cardinality $K_{\mathrm{em}}$ is observed at training time but varies across trains, and the model is expected to operate on previously unseen trains whose values of $K_{\mathrm{em}}$ are unknown a priori.
The catalogue size $M$ is fixed and shared across trains, although individual trains typically exercise only a small subset of the $M$ modes.

\subsection*{Induced partitions, \gls{pdw} streams, and mode hopping}

\Cref{fig:method:pdw_emitter_stream} sketches a short \gls{pdw} stream in chronological order for a single pulse train.
Fill color encodes the train-local emitter index $e_n$; pulses that share a color belong to the same emitter even though they need not be contiguous in time (interleaved emissions from several \gls{radar} platforms are typical).
Each stream cell uses that emitter fill \emph{and} a grayscale pattern for the global operating-mode label $m_n$; the vector label $\vect{x}_i$ inside each cell is drawn in black for legibility.
The right-hand legends separate the two cues: emitter entries are color-only (no pattern), and mode entries are monochrome pattern tiles with black text.
The formal partitions and mode hopping are stated mathematically in the text below.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    x=12mm,
    font=\small,
    pdw/.style={
      draw,
      rounded corners=1.6pt,
      minimum width=8.8mm,
      minimum height=8mm,
      inner sep=1.5pt,
      align=center,
      text=black
    },
    % Stream cells: fill + pattern (path picture paints emitter fill under pattern; plain fill+pattern often hides fill in PDF)
    pdwstreamfill/.style={fill=none},
    >=Latex
  ]
    \foreach \i/\col/\m [count=\n] in {
      1/blue!18/1,2/orange!22/1,3/blue!18/2,4/orange!22/2,5/blue!18/1,
      6/green!20/1,7/orange!22/2,8/blue!18/2,9/green!20/3,10/orange!22/1
    } {
      \ifnum\m=1
        \node[pdw, pdwstreamfill,
          path picture={
            \fill[\col] (path picture bounding box.south west)
              rectangle (path picture bounding box.north east);
            \path[pattern=horizontal lines, pattern color=black!48]
              (path picture bounding box.south west)
              rectangle (path picture bounding box.north east);
          }] (p\n) at (\n, 0) {$\vect{x}_{\i}$};
      \else
        \ifnum\m=2
          \node[pdw, pdwstreamfill,
            path picture={
              \fill[\col] (path picture bounding box.south west)
                rectangle (path picture bounding box.north east);
              \path[pattern=dots, pattern color=black!52]
                (path picture bounding box.south west)
                rectangle (path picture bounding box.north east);
            }] (p\n) at (\n, 0) {$\vect{x}_{\i}$};
        \else
          \node[pdw, pdwstreamfill,
            path picture={
              \fill[\col] (path picture bounding box.south west)
                rectangle (path picture bounding box.north east);
              \path[pattern=north east lines, pattern color=black!48]
                (path picture bounding box.south west)
                rectangle (path picture bounding box.north east);
            }] (p\n) at (\n, 0) {$\vect{x}_{\i}$};
        \fi
      \fi
    }
    \draw[->, thick] (0.35,-0.62) -- ({10+0.65},-0.62)
      node[midway, below=2.5pt, font=\footnotesize] {pulse index $n$ (time order)};
    % Emitter legend: fixed column (swatch east edges aligned; text on a common vertical line)
    \begin{scope}[font=\footnotesize]
      \path let \p{e}=(p10.east) in coordinate (legline) at ({\x{e}+8mm},0);
      % Emitter legend: solid fill only (no pattern); labels in emitter hues
      \foreach \k/\col/\txtcol/\y in {1/blue!18/blue!78!black/0.46,2/orange!22/orange!90!black/0,3/green!20/green!58!black/-0.46} {
        \node[pdw, fill=\col, minimum width=5.2mm, minimum height=4.5mm, anchor=east]
          (legb\k) at ($(legline)+(0,\y)$) {};
        \node[anchor=mid west, text=\txtcol] at ($(legline)+(2.2mm,\y)$) {emitter $\k$};
      }
      % Mode legend: gray + pattern only (no emitter hue); labels in black
      \coordinate (mlegline) at ($(legline)+(26mm,0)$);
      \node[pdw, fill=gray!42, pattern=horizontal lines, pattern color=black!68,
            minimum width=5.2mm, minimum height=4.5mm, anchor=east, draw=black!75]
        (mlegb1) at ($(mlegline)+(0,0.46)$) {};
      \node[anchor=mid west, text=black] at ($(mlegline)+(2.2mm,0.46)$) {mode $1$};
      \node[pdw, fill=gray!42, pattern=dots, pattern color=black!68,
            minimum width=5.2mm, minimum height=4.5mm, anchor=east, draw=black!75]
        (mlegb2) at ($(mlegline)+(0,0)$) {};
      \node[anchor=mid west, text=black] at ($(mlegline)+(2.2mm,0)$) {mode $2$};
      \node[pdw, fill=gray!42, pattern=north east lines, pattern color=black!68,
            minimum width=5.2mm, minimum height=4.5mm, anchor=east, draw=black!75]
        (mlegb3) at ($(mlegline)+(0,-0.46)$) {};
      \node[anchor=mid west, text=black] at ($(mlegline)+(2.2mm,-0.46)$) {mode $3$};
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of a \gls{pdw} stream $\vect{x}_1,\ldots,\vect{x}_N$ for one pulse train, in time order.
    Each cell combines emitter color (fill) and operating-mode pattern; the in-cell vector index is typeset in black.
    The emitter legend uses color-only swatches with matching colored text for the train-local emitter labels; the mode legend uses grayscale pattern swatches with black text for the global mode catalogue.
    Emitter recovery seeks the within-train partition into subsets $\mathcal{E}_k$ (constant $e_n$ on each subset); mode prediction labels each pulse with its global mode $m_n$, while allowing mode hops within a fixed emitter.}
  \label{fig:method:pdw_emitter_stream}
\end{figure}

Within a pulse train, the emitter supervision induces a partition of $\{1,\ldots,N\}$ into emitter-coherent index sets
\begin{equation}\label{eq:method:emitter_partition}
  \mathcal{E}_k = \{\, n \in \{1,\ldots,N\} : e_n = k \,\},
  \qquad
  k \in \{1,\ldots,K_{\mathrm{em}}\},
\end{equation}
so that $\vect{x}_n$ and $\vect{x}_{n'}$ belong to the same emitter-coherent subset if and only if $e_n = e_{n'}$.
Each $\mathcal{E}_k$ is the preimage of label $k$ under the map $n \mapsto e_n$, and $\{\mathcal{E}_1,\ldots,\mathcal{E}_{K_{\mathrm{em}}}\}$ is pairwise disjoint with union $\{1,\ldots,N\}$.
Because emitter symbols are not shared across trains, this equivalence relation is defined only within each train; the recovery problem at inference time is to reconstruct it from $\mat{X}$ alone.

The global mode catalogue induces a second partition,
\begin{equation}\label{eq:method:mode_partition}
  \mathcal{M}_\ell = \{\, n \in \{1,\ldots,N\} : m_n = \ell \,\},
  \qquad
  \ell \in \{1,\ldots,M\},
\end{equation}
which groups pulses that share the same operating mode regardless of which physical emitter produced them.
The two partitions are coupled because $(e_n,m_n)$ is generated jointly: restricting to a fixed emitter $k$, the induced set of mode labels
\begin{equation}\label{eq:method:modes_per_emitter}
  \mathcal{L}_k = \{\, m_n : n \in \mathcal{E}_k \,\} \subseteq \{1,\ldots,M\}
\end{equation}
has cardinality $L_k = \lvert \mathcal{L}_k \rvert$ bounded by the number of distinct waveform programmes that emitter $k$ exercises.
In many \gls{radar} systems, mode hopping switches among a small subset of the catalogue, so $L_k$ is modest---often on the order of single-digit integers---even though the temporal pattern $(m_n)_{n \in \mathcal{E}_k}$ is not constant.
The supervised learning problem is then to fit a model whose outputs, on previously unseen trains, recover both partitions up to arbitrary relabellings of the cluster symbols, using the dual embeddings in \cref{eq:method:dual_embeddings}: for the emitter partition the relabelling acts within each train, while for the mode partition the relabelling acts globally and may introduce additional indices when previously unseen modulations are present.

\subsection*{Desired outputs: dual embeddings and discrete labels}

The parametric model considered in this thesis maps an input pulse train to two parallel trajectories in Euclidean embedding spaces,
\begin{equation}\label{eq:method:dual_embeddings}
  \mat{Z}^{\mathrm{em}} = (\vect{z}^{\mathrm{em}}_1,\ldots,\vect{z}^{\mathrm{em}}_N),
  \quad
  \vect{z}^{\mathrm{em}}_n \in \R^{p},
  \qquad
  \mat{Z}^{\mathrm{md}} = (\vect{z}^{\mathrm{md}}_1,\ldots,\vect{z}^{\mathrm{md}}_N),
  \quad
  \vect{z}^{\mathrm{md}}_n \in \R^{q},
\end{equation}
with dimensions $p$ and $q$ fixed by the architecture.

The two trajectories play different roles dictated by the label semantics described above, but in both cases the supervision shapes embedding \emph{geometry} so that downstream discretisation is done by clustering rather than by a fixed-class softmax.
Emitter-oriented vectors $\vect{z}^{\mathrm{em}}_n$ are trained so that, within each pulse train, embeddings with the same emitter label cluster tightly in $\R^{p}$ while embeddings with different emitter labels are pushed apart.
Because the emitter symbol is train-local, the corresponding supervision is a within-train metric-learning signal rather than a global classifier, and discrete emitter assignments
\begin{equation}\label{eq:method:emitter_predictor}
  \hat{e}_n \in \{1,\ldots,\hat{K}_{\mathrm{em}}\},
  \qquad
  n \in \{1,\ldots,N\},
\end{equation}
are obtained \emph{post hoc} by clustering $\{\vect{z}^{\mathrm{em}}_n\}_{n=1}^{N}$ within each train; the inferred count $\hat{K}_{\mathrm{em}}$ is either selected by the clustering algorithm or supplied from external knowledge.
Their main deliverable is a per-pulse emitter index suitable for downstream \gls{elint} workflows such as emitter counting and localisation.

Mode-oriented vectors $\vect{z}^{\mathrm{md}}_n$ are trained against the global mode labels analogously: embeddings of pulses sharing the same mode in $\{1,\ldots,M\}$ are pulled together while embeddings of pulses with different modes are pushed apart, this time with positives drawn from the entire batch rather than only from within a single train.
Discrete mode assignments
\begin{equation}\label{eq:method:mode_predictor}
  \hat{m}_n \in \{1,\ldots,\hat{M}\},
  \qquad
  n \in \{1,\ldots,N\},
\end{equation}
are obtained \emph{post hoc} by clustering $\{\vect{z}^{\mathrm{md}}_n\}_{n=1}^{N}$, optionally aided by the cluster centroids of the training-time modes when an alignment to the catalogue $\{1,\ldots,M\}$ is desired.
The inferred catalogue size $\hat{M}$ may exceed $M$ when modulation types that were absent or held out during training are exercised at inference; in that regime, additional clusters are expected to emerge as evidence of previously unseen modes rather than to be forced into existing catalogue entries.
The embedding additionally aims to summarise mode-dependent waveform behaviour so that operating modes can be inspected, compared, or matched against external waveform libraries.

Architecturally, both streams are produced from a shared encoder $\Enc$ with task-specific heads (see \cref{sec:method:arch}); in deployment, the per-pulse deliverables are the discrete pairs $(\hat{e}_n,\hat{m}_n)$, while the continuous embeddings remain available for inspection, prototype matching, and downstream tasks that benefit from a geometric rather than purely categorical mode representation.

\subsection*{Learning objective}

The optimisation problem combines two supervised metric-learning losses sharing the same functional form but operating on different scopes: (i) a within-train term that uses the emitter labels $\{e_n\}$ to shape $\mat{Z}^{\mathrm{em}}$, with positives drawn only from the same pulse train; and (ii) a global term that uses the mode labels $\{m_n\}$ to shape $\mat{Z}^{\mathrm{md}}$, with positives drawn from anywhere in the batch.
The two terms exploit the asymmetric label semantics: emitter labels enter the loss only through within-train comparisons because the symbols are not shared across trains, while mode labels are used identically across trains because the catalogue is global.
Neither term commits to a fixed output cardinality at inference; downstream cluster counts $\hat{K}_{\mathrm{em}}$ and $\hat{M}$ are read from the post-hoc clustering of \cref{eq:method:emitter_predictor,eq:method:mode_predictor} and may differ from the training-time cardinalities.
The concrete loss terms and training protocol are given in \cref{sec:method:loss,sec:method:training}.
