\section{Data Representation and Preprocessing}\label{sec:method:data}

\Gls{elint} data are difficult to obtain at scale, and operational recordings are often sensitive.
This work therefore relies on synthetic pulse trains from a proprietary scenario simulator at Saab, which yields controlled scenarios and full supervision of emitter identity and operating mode for evaluation, while the learning objectives remain unsupervised with respect to those labels.
Earlier KTH degree projects hosted at Saab document synthetic radar-\gls{ml} pipelines at a comparable level of detail, including scenario-driven pulse generation and explicit noise models \parencite{jonsson2023supcon,svensson2022priemitter}.

\subsection{Scenario simulation and pulse records}\label{sec:method:data:sim}

The simulator produces \emph{scenarios}: time-ordered pulse-level timelines for one or more notional emitters and a receiving platform.
Each pulse is described by the continuous fields that enter the \gls{pdw} representation used here (including \gls{toa}, \gls{rf}, pulse width, and amplitude on a logarithmic scale), together with discrete labels for emitter identity and operating mode that exist only for evaluation.
\textcite{jonsson2023supcon} describes the unclassified OpenModesty tool in the same lineage: flight paths and emitter placements define a geometry, pulses are emitted and intercepted one at a time, and the simulator can apply modulations and controlled perturbations consistent with notional radar behavior.
The present implementation is proprietary; the important methodological point is that every pulse in the export carries consistent timing and parameter fields so that windows formed below correspond to reproducible segments of a simulated electromagnetic timeline.

Scenarios differ in the number of emitters, dwell in each mode, and simulated horizon so that training covers diverse traffic densities; exact counts and split assignments are deferred to \cref{ch:experiments}.

\subsection{Optional receiver-side degradation}\label{sec:method:data:noise}

Even when emitter physics are ideal in the simulator, downstream \gls{esm} processing introduces structured errors.
\textcite{jonsson2023supcon} distinguishes \emph{dropped pulses}, where a transmitted pulse is not registered by the receiver chain, from \emph{spurious pulses}, where extra pulses appear in a track because deinterleaving or detection stages attach energy to the wrong pulse train.
\textcite{svensson2022priemitter} likewise emphasizes controlled drop, spurious, and additive Gaussian regimes when studying \gls{pri} sequences.
When the experimental programme applies such degradations, they are implemented as stochastic edits on the ordered pulse list \emph{before} windowing: each pulse may be removed with a scenario- or split-specific drop probability, and spurious insertions may be drawn by copying and perturbing existing pulses or by drawing alternative parameter vectors from a bounded rule set.
Discrete grids of drop and spurious rates, rather than only single operating points, simplify reporting across noise sweeps in \cref{ch:experiments}.

\begin{algorithm}[t]
  \caption{Optional stochastic degradation of an ordered pulse list (conceptual).}\label{alg:method:pulse_degradation}
  \begin{algorithmic}[1]
    \Require Ordered pulses $\{p_1,\ldots,p_N\}$, drop probability $\rho_{\mathrm{drop}} \in [0,1)$, spurious probability $\rho_{\mathrm{spu}} \in [0,1)$
    \Ensure Degraded list of pulses preserving time order where applicable
    \State $\mathcal{L} \gets \mathrm{copy}(\{p_1,\ldots,p_N\})$
    \For{each pulse $q$ in $\mathcal{L}$ scanned in forward time order}
      \If{uniform draw $u < \rho_{\mathrm{drop}}$}
        \State remove $q$ from $\mathcal{L}$
      \EndIf
    \EndFor
    \For{each remaining pulse $q$ in $\mathcal{L}$}
      \If{uniform draw $u' < \rho_{\mathrm{spu}}$}
        \State insert a spurious pulse after $q$ using the chosen perturbation rule (duplicate with noise or independent draw)
      \EndIf
    \EndFor
  \end{algorithmic}
\end{algorithm}

\cref{alg:method:pulse_degradation} is only a bookkeeping template: concrete probabilities, conditioning on track boundaries, and whether noise is re-drawn per window or fixed per scenario appear in the experiment log together with the simulator export manifest.

\subsection{Windowing and context length}\label{sec:method:data:window}

The encoder operates on fixed-length windows because standard self-attention builds an $L \times L$ compatibility matrix between tokens at each layer, so time and memory scale as $\mathcal{O}(L^2)$ in the window length $L$ at fixed model width, repeated across depth.
Let $L$ denote the window length (in pulses) fed to the model; experiments use $L = 1024$.
Each scenario is partitioned into contiguous windows of length $L$.
If the remainder after the last full window is shorter than $L$, it is dropped in the present pipeline; padding with an attention mask would retain the tail without changing the architecture and is left as a straightforward extension.

Training windows are extracted with a stride of $L/2$, so consecutive windows overlap by half their length.
This increases the number of training pulses seen per scenario and reduces sensitivity to the particular alignment of window boundaries relative to mode transitions.
Validation and test windows use stride $L$ (no overlap), so that metrics are not inflated by near-duplicate windows.

Interleaved traffic and heterogeneous dwell times imply that pulse counts per emitter and per operating mode vary widely inside a window, and the effective \gls{pri} structure can make the number of pulses per emitter differ across the corpus even when scenario durations match \parencite{jonsson2023supcon}.
Some windows therefore contain few or no pulses from a rare emitter or a short-lived mode, even when that label is frequent in the parent scenario, and dropped tails can remove trailing segments entirely.
These effects induce a form of \emph{label incidence imbalance} at the window level that is distinct from corpus-level class histograms and should be read alongside the reporting conventions in \cref{ch:experiments}.

\subsection{Per-feature scaling}\label{sec:method:data:norm}

Amplitude is already provided on a logarithmic scale (decibels).
Continuous fields are transformed before windowing so that statistics are computed on the training split only and then applied to validation and test data.
The log-amplitude, carrier frequency, and pulse width are standardised to zero mean and unit variance using the training-set mean and standard deviation of each field, so that channels measured in different physical units contribute comparably to dot products in the encoder \parencite{jonsson2023supcon}.
Each \gls{toa} is first min--max scaled to $[0,1]$ using training-set bounds; within every window, all rescaled \gls{toa} values are shifted by subtracting the \gls{toa} of the first pulse so that the window starts at time zero.
This removes absolute clock offset while preserving intra-window timing structure.

% TODO: Encoding of any categorical PDW fields once the feature list is fixed.

\subsection{Augmentations for contrastive views}\label{sec:method:data:aug}

Contrastive objectives in \cref{sec:method:loss} consume two stochastic views of each training window.
Those views combine the deterministic normalisation above with additional pulse-level or window-level perturbations (for example timing jitter, amplitude scaling, small frequency offsets, or pulse dropout) so that invariances reflect both simulator noise and plausible analogue front-end variability.
The exact augmentation set and sampling ranges are part of the training configuration and are listed with other hyperparameters in \cref{sec:exp:hparams}.

Windows are grouped into mini-batches for optimisation; batch size and per-epoch shuffling of windows are stated in \cref{sec:exp:hparams}.
