\section{Transformer Encoder Architecture}\label{sec:method:arch}

The encoder $\Enc$ takes the form of a transformer-encoder backbone \parencite{vaswani2017attention} with a shared trunk followed by two task-specific branches.
This layout mirrors the dual-embedding structure introduced in \cref{sec:method:formulation}: the trunk produces a single contextualised representation of the input window, which the branches specialise into emitter-oriented and mode-oriented embeddings.
% TODO: include figures/method/architecture.pdf once the schematic is drawn.

\subsection*{Input embedding}

Each \gls{pdw} vector $\vect{x}_n \in \R^{d}$, normalised as described in \cref{sec:method:data}, is mapped to a token embedding through an affine projection
\begin{equation}\label{eq:method:input_embed}
  \vect{h}^{(0)}_n = \mat{W}_{\mathrm{in}}\vect{x}_n + \vect{b}_{\mathrm{in}},
  \qquad
  \vect{h}^{(0)}_n \in \R^{d_{\mathrm{model}}},
\end{equation}
where $\mat{W}_{\mathrm{in}} \in \R^{d_{\mathrm{model}} \times d}$ and $d_{\mathrm{model}}$ is the trunk width specified below.
No positional encoding is added to $\vect{h}^{(0)}_n$.
The temporal ordering of pulses is already carried by the \gls{toa} entry of $\vect{x}_n$, which makes a separate position signal redundant; consistent with this argument, \textcite{gunn2025deinterleaving} report that adding sinusoidal or learnable positional encodings to a transformer deinterleaver yields a small but reproducible drop in performance, attributed to the resulting redundancy with the \gls{toa} feature.

\subsection*{Shared trunk}

The token sequence $(\vect{h}^{(0)}_1,\ldots,\vect{h}^{(0)}_L)$ is processed by a stack of $N_{\mathrm{sh}}$ standard transformer-encoder layers, each comprising a \gls{mha} sub-layer and a position-wise \gls{ffn} sub-layer with residual connections and layer normalisation around both.
Self-attention is unmasked, so every pulse can attend to every other pulse in the window.
The trunk operates with hidden dimension $d_{\mathrm{model}} = 256$, $h_{\mathrm{sh}} = 8$ attention heads (so that each head has dimension $32$), and an inner \gls{ffn} dimension of $d_{\mathrm{ff,sh}} = 2048$; $N_{\mathrm{sh}} = 6$ layers are used.
Its output is a contextualised token sequence $\mat{H}^{\mathrm{sh}} = (\vect{h}^{\mathrm{sh}}_1,\ldots,\vect{h}^{\mathrm{sh}}_L)$ with $\vect{h}^{\mathrm{sh}}_n \in \R^{256}$, in which information has been mixed across the full window.

\subsection*{Task-specific branches}

The trunk output feeds two parallel branches that do not share parameters, producing the dual embeddings of \cref{eq:method:dual_embeddings}.
Each branch begins with an affine projection that reduces the hidden width from $d_{\mathrm{model}} = 256$ to $d_{\mathrm{br}} = 128$, after which $N_{\mathrm{br}} = 2$ transformer-encoder layers of the same form as the trunk are applied.
Within a branch, \gls{mha} uses $h_{\mathrm{br}} = 4$ heads (again with per-head dimension $32$) and the \gls{ffn} inner dimension is $d_{\mathrm{ff,br}} = 1024$.
The reduced width and depth reflect a deliberate asymmetry: the trunk is intended to carry the heavy lifting of contextual mixing, while the branches only need to specialise the representation to either emitter identity or operating mode.
Write the resulting token sequences as $\mat{H}^{\mathrm{em}} = (\vect{h}^{\mathrm{em}}_1,\ldots,\vect{h}^{\mathrm{em}}_L)$ and $\mat{H}^{\mathrm{md}} = (\vect{h}^{\mathrm{md}}_1,\ldots,\vect{h}^{\mathrm{md}}_L)$ with $\vect{h}^{\mathrm{em}}_n, \vect{h}^{\mathrm{md}}_n \in \R^{128}$.

\subsection*{Projection heads}

Each branch terminates in a linear projection head that maps every token to its respective embedding space,
\begin{equation}\label{eq:method:proj_heads}
  \vect{z}^{\mathrm{em}}_n = \mat{W}_{\mathrm{em}}\vect{h}^{\mathrm{em}}_n + \vect{b}_{\mathrm{em}},
  \qquad
  \vect{z}^{\mathrm{md}}_n = \mat{W}_{\mathrm{md}}\vect{h}^{\mathrm{md}}_n + \vect{b}_{\mathrm{md}},
\end{equation}
with $\vect{z}^{\mathrm{em}}_n \in \R^{p}$ and $\vect{z}^{\mathrm{md}}_n \in \R^{q}$ for $p = 8$ and $q = 16$.
The mode space is assigned the higher dimensionality because, conditional on emitter identity, mode structure is expected to require a richer geometry to separate subtle waveform regimes, whereas emitter membership induces a coarser partition over the same pulses and is well served by a more compact embedding.

\subsection*{Per-window clustering with \gls{hdbscan}}

The projection heads in \cref{eq:method:proj_heads} map each pulse to a continuous vector, whereas the latent emitter and mode assignments in \cref{sec:method:formulation} are discrete.
\Gls{hdbscan} \parencite{campello2015hierarchical,mcinnes2017hdbscan} is applied independently to the two embedding trajectories within every analysis window: the $L$ points $\{\vect{z}^{\mathrm{em}}_n\}_{n=1}^L$ and $\{\vect{z}^{\mathrm{md}}_n\}_{n=1}^L$ each form a separate clustering instance on the corresponding branch.
Each window is thus a self-contained pulse train segment in embedding space, consistent with the windowing of \cref{sec:method:data}, and the effective cardinalities of the inferred emitter and mode partitions follow the density hierarchy instead of fixing $K_{\mathrm{em}}$ or $K_{\mathrm{md}}$ a priori.

As defined in the framework of \textcite{campello2015hierarchical}, \gls{hdbscan} replaces the single global density threshold of DBSCAN \parencite{ester1996density} with a hierarchy of density estimates: nested level sets are summarised in a condensed tree, and a flat clustering is read off by favouring groups that persist as coherent peaks rather than artefacts of a particular threshold.
The same construction supports an explicit noise category for points that are not assigned to any cluster leaf retained after the hierarchy is flattened \parencite{campello2015hierarchical}.
The computations use the reference library described by \textcite{mcinnes2017hdbscan}.

The collection of all linear maps in \cref{eq:method:input_embed,eq:method:proj_heads} together with the attention and \gls{ffn} weights of the trunk and branches constitutes the parameter vector $\bm{\theta}$ that the loss in \cref{sec:method:loss} optimises.
